---
title: "Course Project - Practical Machine Learning"
author: "Ange Liu"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: true 
      smooth_scroll: true
    theme: "lumen"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse);library(caret)
library(knitr);library(DT);library(randomForest)
```

## Background and Objective

6 participants were asked to perform barbell lifts in 6 (1 correct and 5 incorrect) different ways. Using the data from accelerometers on the belt, forearm, arm, and dumbell of the 6 participants during the exercise, we will try to predict the manner of the excercise performed.

## Getting and exploring the data

First, we download the data from the following links and save them:

```{r message=FALSE, cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
              "./data/pml-training.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
              "./data/pml-testing.csv")
```

We will load only the **training** data for now, leaving the **test** data aside.  

```{r}
training <- read.csv("./data/pml-training.csv", stringsAsFactors = FALSE,
                     ## attn: several different NA strings
                     na.strings = c("", "#DIV/0!", NA_character_))
```

Since there are a lot of variables in the data set (`r ncol(training)`), let's first explore them before fitting the models.

```{r}
## create a dataframe and store the characteristics of the variables
ref.columns <- tibble(
      ## column names
      COLNAME = colnames(training),
      ## column class
      CLASS = map_chr(training, class) %>% unname(),
      ## how many NAs in each column
      N_NA = map_int(training, ~ is.na(.x) %>% sum())) %>% 
  ## what is the percentage of NAs for each column
  mutate(PERCENT_NA = round(N_NA / nrow(training), 3))

## print the table nicely
datatable(ref.columns, filter = "top")
```

## Data Preprocessing

### Data Cleaning

Looking at the table, we can see many of the variables have a significant percentage of missing values. We will clean up the data by setting a threshold (e.g. 90%) and exclude those with higher percentage of missing values from the model fitting. In addition, we will also exclude those variables that are not the measurements from the accelerometers.

The variables that we will end up keeping are:
```{r}
InCols <- ref.columns %>% 
  filter(PERCENT_NA < 0.9, 
         !str_detect(COLNAME, "X|user_name|timestamp|window")) %>% 
  pull(COLNAME) %>% 
  print()
```

### Data Partitioning

Since the **test** data from the source will be used for the *Prediction Quiz*, we will have to withhold part of the **training** data from the model and use them as **testing** data to evaluate the models.

```{r}
set.seed(202)
InTrain <- createDataPartition(y = training$classe, p = 0.75, list = FALSE)
part_train <- training[InTrain, InCols]
part_test <- training[-InTrain,] %>% mutate(classe = factor(classe))
```

## Model Building

For this project, we will try the following methods using the partitioned training data and compare the accuracy of the models using the partitioned test data.

* Classification and Regressio Trees (CART)
* Gradient Boosting Model (GBM)
* Random Forest (RF)

### Fit and save the models

*CART Model*
```{r, cache = TRUE}
## fit the CART model
model_rpart <- train(classe ~ ., data = part_train, method = "rpart")

## cache the fitted model
saveRDS(model_rpart, "./model/model_rpart.rds")
```

*GMB*
```{r, cache = TRUE, results = "hide"}
## fit the GBM
model_gbm <- train(classe ~ ., data = part_train, method = "gbm")

## cache the fitted model
saveRDS(model_gbm, "./model/model_gbm.rds")
```

*RF*
```{r, cache = TRUE, eval=FALSE}
## fit the rf model
model_rf <- train(classe ~ ., data = part_train, method = "rf")

## cache the fitted model
saveRDS(model_rf, "./model/model_rf.rds")
```

### Model selection

Using the withheld partition, we will predict the manner of excercise using each model and compare the predicted values with the true ones.

```{r}
## load the saved model
model_rpart <- readRDS("./model/model_rpart.rds")
model_gbm <- readRDS("./model/model_gbm.rds")
model_rf <- readRDS("./model/model_rf.rds")

## predict the class
predict_rpart <- predict(model_rpart, newdata = part_test)
predict_gbm <- predict(model_gbm, newdata = part_test)
predict_rf <- predict(model_rf, newdata = part_test)

## get the confusion matrix for the models
cfm_rpart <- confusionMatrix(predict_rpart, part_test$classe)
cfm_gbm <- confusionMatrix(predict_gbm, part_test$classe)
cfm_rf <- confusionMatrix(predict_rf, part_test$classe)
```

The model accuracies are the following. The RF model is the one with the highest accuracy.

```{r}
tibble(MODEL = c("CART", "GBM", "RF"),
       ACCURACY = c(cfm_rpart$overall["Accuracy"],
                        cfm_gbm$overall["Accuracy"],
                        cfm_rf$overall["Accuracy"]))
```

We can take a look at the variable importance of the wining RF model:

```{r, message = FALSE}
varImp(model_rf)
```

## Prediction Quiz

Now that we have settled on a model, we can proceed to predict the manner of excercise in the **test** data from the source.

```{r}
## load the downloaded data
test <- read.csv("./data/pml-testing.csv", stringsAsFactors = FALSE,
                    ## attn: several different NA strings
                    na.strings = c("", "#DIV/0!", NA_character_))

## make the prediction
predict_quiz <- predict(model_rf, newdata = test)

## the prediction results are the following
tibble(PROBLEM_ID = test$problem_id, PREDICTED_VALUE = predict_quiz)
```

