---
title: "Statistical Inference Project - Part 2 Inferential Data Analysis"
author: "ALIU"
date: "September 18, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Load the ToothGrowth data and Basic exploratory data analyses

Load the data “ToothGrowth” into R and take a look at the structure using the 
"str" function. 

```{r}
data(ToothGrowth)
str(ToothGrowth)
```

To explore the relationship between “len” and “dose”, we can plot the two 
variables in an x-y plot, given the type of “supp”.

```{r, fig.height= 2.5, fig.width= 5}
g <- ggplot(ToothGrowth, aes(x=dose, y=len, colour=supp)) + 
  geom_point(shape=1) + geom_smooth(method=lm)
g
```

## Basic summary of the data
From the outputs, we can see that there are 3 variables and 60 
observations in the data set, with the variables "len" and "dose" as numeric 
variables and "supp" as a factor variable of two levels. From

the plot, we can see that as the dose increases, the lent of the tooth tend to 
be longer as well and this trend is true for both supplement categories. At the 
same dose level, the supplement OJ seems to have a greater effect on tooth 
length, although, as the dose increases, the difference between OJ and VC decreases

## Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.

### Tooth growth by supplement
To compare the tooth growth by supplement, we can Perform a hypothesis test with the null hypothesis: OJ and VC do not have a statistically significant difference in their effect on tooth growth. The results are as follows:

```{r}
test1 <- t.test(len ~ supp,data=ToothGrowth)
test1
```

The p-value of  `r test1$p.value` is suggests that we do not have sufficient evidence to reject the null hypothesis that OJ and VC do not have a statistically significant effect on tooth growth. The same conclusion can also be drawn by interpreting the confidence interval of `r test1$conf.int` for mean(OJ)-mean(VC), which contains zero.

### Tooth growth by dose
Similiarly, to compare the tooth growth by dose, we can Perform a hypothesis test with the null hypothesis: dose do not have a statistically significant difference in their effect on tooth growth. For this analysis, we will ignore the type of supplement and subset the three dosage levels into groups for the hypothesis test. For simplicity, only the p-values and confidence intervals of each test are shown. The results are as follows:

```{r}
dose12 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
dose23 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
dose13 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))

test2a <- t.test(len ~ dose, data=dose12)
test2b <- t.test(len ~ dose, data=dose13)
test2c <- t.test(len ~ dose, data=dose23)

test2 <- data.frame(p.value = c(test2a$p.value, test2b$p.value, test2c$p.value),
                    conf.int.lower = c(test2a$conf.int[1],
                                                  test2b$conf.int[1],
                                                  test2c$conf.int[1]),
                    conf.int.upper = c(test2a$conf.int[2],
                                                  test2b$conf.int[2],
                                                  test2c$conf.int[2]),
                    row.names = c("Test 2A", "Test 2B", "Test 2C"))

test2
```

With p-values all below 0.05, and the confidence intervals not containing zero, we can reject the null hypotheses and conclude that there is a significant difference in tooth length between dosages among all three levels.

### Comparison of tooth growth considering both factors
To compare tooth growth considring both dose and supplement, we subset the data into groups by the dosage levels and perform hyphothesis tests for the supplement types at each dosage level.

```{r}
dose05 <- subset(ToothGrowth, dose == 0.5); test3a <- t.test(len ~ supp, data=dose05)
dose10 <- subset(ToothGrowth, dose == 1.0); test3b <- t.test(len ~ supp, data=dose10)
dose20 <- subset(ToothGrowth, dose == 2.0); test3c <- t.test(len ~ supp, data=dose20)

test3 <- data.frame(p.value = c(test3a$p.value, test3b$p.value, test3c$p.value),
                    conf.int.lower = c(test3a$conf.int[1],
                                                  test3b$conf.int[1],
                                                  test3c$conf.int[1]),
                    conf.int.upper = c(test3a$conf.int[2],
                                                  test3b$conf.int[2],
                                                  test3c$conf.int[2]),
                    row.names = c("Test 3A", "Test 3B", "Test 3C"))

test3

```

Based on the p-values and confidence intervals, we can reject the null hypotheses and conclude that OJ is significantly more effective than VC for tooth growth at dose levels 0.5 and 1.0. However, when dose level reaches 2.0, we cannot reject the null hypothesis and therefore have to conclude that the difference in effect on tooth growth between OJ and VC is not statistically significant. 

## Conclusions and the assumptions needed

**Conclusions**

- When ignoring dose levels, there is no statistically significant difference in the tooth growth between the supplement types.
- When ignoring the supplement types, there is a statistically significant difference in tooth length between the dose levels, with higher doses resulting in longer teeth.
- When considering both factors:
    - At the dosage levels 0.5 and 1.0, OJ has a statistically significant effect on tooth length comparing to VC
    - However, as the dosage increases to 2.0, the difference is no longer significant

**Assumptions:**
In the tests and analysis performed, we assume that the variances between the separate populations tested are different (used the default var.equal = FALSE for all the t tests). Also, we assume that the populations are independent and random.